The research leading to these results has received funding from the European
Community's Seventh Framework
Programme (FP7/2007--2013) under grant agreement n~211743. \par 
We thank the whole ET science team for the excellent activities 
during the design study phase and for the great support in producing this document.\par
We also thank the German Centre for Quantum Engineering and Space-Time Research, QUEST, for support.\par
This work is part of the research programme of the Foundation for Fundamental Research
 on Matter (FOM), which is financially supported by the Netherlands Organisation for
  Scientific Research (NWO).\par
  All the seismic measurements have been made possible with the dedication and help from a vast number of locals. To all the people that have provided their expertise, time and assistance we would like to extend a sincere thank you. Due to the extent of the research it is impractical to list all those people but we would like to thank a few people in particular. For fruitful discussions and advice as well as facilitating measurements at the Heimansgroeve we thank Reinoud Sleeman and his colleagues from the KNMI. We also acknowledge their help in obtaining data from the Virtual European Broadband Seismograph Network (VEBSN). The measurements in Hungary were generously supported by members of the Research Institute for Particle and Nuclear Physics, Budapest, and the mining company MECSEK-\"OKO. For assistance during the measurements in Romania we are grateful to Romul Mircea Margineanu and the Salrom mining company. The Frejus tunnel measurements were kindly facilitated by Michel Zampaolo of the Laboratoire Souterrain de Modane. We thank Alessandro Bettini and Jose Jimenez of the Canfranc Underground Laboratory for their help during measurements in Spain. We are grateful to the colleagues from the University ``Sapienza'' and the miners of IGEA spa, for access to the Sardinian site. We would also like to thank Eugenio Coccia and Benedetto Gallese for their help during measurements at the Gran Sasso underground laboratory. We are very grateful to Rudolf Widmer-Schnidrig, Thomas Forbinger and Walter Zuern of the Black Forest Observatory for the interesting discussions and access to their laboratory. We acknowledge Guido Nuijton for his thoughts on sites in Finland. For measurements carried out at the CLIO site in Japan we thank Uchiyama Takashi, Osamu Miyakawa and Miyoki Shinji for their external help. Ricardo de Salvo and friends at the Italkali mine in Sicilie kindly assisted the measurements there. Finally we would like to acknowledge the help of EURIDICE staff members, Philippe van Marcke and Kris Moerkens, during measurements in Belgium.\par
We thank R.Prix for helping with Fig.~\ref{fig-CW-known}.