\subsection{ Cryogenic compatibility of materials for Superattenuator construction}
\label{sec:cryo_filters}

In this appendix some guide lines to select materials and components for the construction of a \emph{Superattenuator} working in a cryogenic environment are reported. Doing this it will help in the understanding process of the interface problems linked to the presence of a cryogenic payload suspended from an anti-seismic system. During our investigation, based on data coming from literature and on some lab tests performed by Virgo collaboration, no preliminary hypothesis on the operating temperature has been done keeping the maximum flexibility. Even if some construction materials (stainless steel, $OFHC$ copper, etc.) can be used down to liquid helium temperature this does not represents a limiting factor of our investigation. However, for each element of the chain, it is necessary to perform a careful characterisation test of the thermal expansion coefficient. The presence of very sophisticated optics, indeed, requests a particular attention to avoid the environment contamination that needs to be carefully controlled. The conclusions of our activity, summarised in the following, are discussed in detail in \cite{Poggiani2009}.
\begin{itemize}
\item $Actuators$: the position of different elements of the suspension system, especially those ones close to the optical payload, must be moved for the interferometer control. Some commercial stepping motors designed for Ultra High Vacuum ($UHV$) and cryogenic applications are available, but they should be tested and qualified by a dedicated characterisation measurement performed in the standard working conditions. An AML motor has been tested successfully at Virgo down to a few Kelvin [$F. Frasconi$, $Private$ $communication$]. Cryogenic lubricants should also be selected, avoiding those ones having chemical composition based on hydrocarbon and fluorine for possible contamination issues.
\item $Cabling$: the great variety of control systems and sensors in Einstein Telescope will require a large number of electrical connections, that should have low thermal conductivity (to reduce heat exchange), low electrical resistivity (to reduce dissipation), high flexibility (to reduce noise injection) and low contamination. An optimization of conductors and insulator dimensions as a function of current and operating temperature is required. As detailed in \cite{Poggiani2009}, some low out-gassing samples tested in Virgo can be used at low temperatures: Pyre-Ml insulated wires, alumina insulated wires, Gore-Tex ribbons. While $Kapton$ and $PTFE$ insulation have been already tested for satellite applications.
\item $Adhesives$: during the construction of the interferometer there will certainly be the need to bond several elements of Einstein Telescope. For this reason some adhesives could be selected and used. They must be able to withstand thermal cycling and match the thermal expansion coefficients of the bonded elements. The adhesives properties are strongly dependent on the specific batch and on the curing recipe (time and temperature). As shown in \cite{Poggiani2009} several solutions are available for testing them down to liquid helium temperature, including epoxy and ceramic adhesives. Even if some of them have been tested on optical surfaces, the bonding strength, the thermal expansion coefficient, ageing and contamination issues on long time scales must be measured by a dedicated research activity devoted to a precise characterisation of each adhesive.
\item $Magnets$: the properties of magnetic materials are strongly dependent on the temperature. Ferrite and Alnico materials are not suited below 200 K because they change magnetic properties. Neodymium-Iron-Boron magnets show an increasing magnetic field going from room temperature down to 140-120 K while decreasing its value at lower temperature. Samarium-Cobalt magnets show a magnetic field slowly decreasing with decreasing temperature and they have been successfully used down to liquid helium temperature. A critical point is the "Barkhausen noise" that we are studying in a dedicated $R\&D$ activity for a precise characterisation, as reported in the coil -magnet actuator section. 
\end{itemize}
As mentioned in the suspension chapter, another important topic deals with the creep noise of suspension elastic elements generated by the high stress applied. It has been already shown in the same section  that the noise due to the dislocations motion and to their avalanche formation are typical events triggered by the temperature increasing, important only in the last stage(s) of the \emph{Superattenuator}. The natural question is: what is the situation of this micro-glitch noise in a cryogenic environment? Even if creep is associated to an increment of the environmental temperature, cryogenic creep is not necessarily negligible, as observed for $OFHC$ copper. Data from literature on this subject and relative to steels as reported in \cite{Poggiani2010} are insufficient and scattered. Since most part of previous studies investigated transient creep, it is suggested to perform very long term tests before using this material in a cryogenic environment. In \cite{Poggiani2010} typical apparatus to measure creep in a cryogenic environment are outlined. An $R\&D$ program is required only if in the final design of Einstein Telescope some elastic elements will operate at low temperature. 

\input{Sec_Suspensions/Gas_spring}
\input{Sec_Suspensions/LIGO_suspension}


